State of the Art in Scientific Workflow Systems

### 2014-06-01 paper.tex
separate desc

\paragraph{Taverna/scufl} The Taverna Workflow Management System
\cite{taverna2013} is an open source scientific dataflow manager
developed by the mygrid~consortium in the UK.
%
Taverna includes a GUI-based rich-client workbench for workflow design and a
workflow enactment runtime. Taverna can enact workflows expressed in the scufl
language. Scufl is a near Turing complete dataflow language with well-defined
semantics \cite{turi-missier-etal:2007,glatard-montagnat:2008}.

A Taverna workflow consists of a collection of processors connected by data and
coordination links, which establish a dependency between the output(s) of a
processor and the input(s) of another. The processors are of different kinds
depending on the application code to be invoked: typically being services, with
input and output \textit{ports} that correspond to the operation's invocations
and response sequences defined in the service`s interface expressed as REST or WSDL.
Java classes and local scripts can be used as workflow processors, as long as
they expose their signature as input and output ports as required by the model.
The engine orchestrates the execution of the processors in a way that is
consistent with the dependencies, and manages the flow of data through the
processors' ports. The overall workflow execution is data-driven, with a
\emph{push} model: a processor's execution is started as soon as all of its
inputs are available. Taverna is a flexible and extendible workflow environment
\cite{missier-soiland-reyes-etal:2010,sroka2009a} in that extending the
engine's functionality is facilitated through a Service Provider Interface
(SPI). The Taverna engine implements a sophisticated mechanism for enactment of
parallel workflows. These mechanisms manifest in the form of complex pipelines
and an advanced superpipelined enactment of workflows. Such a mechanism induces
multiple downstream workflow processors running in parallel for a dataset. In
the case of superpipelines, multiple pipeline instances run for multi-datasets
providing a high degree of parallelism limited only by the availability of
underlying resources. Taverna is not natively interfaced with external
computational infrastructure but the same can be achieved with the help of
extending the engine through one of its
SPIs~\cite{maheshwari-missier-etal:2009}. In a crucial development, recent
implementations of Taverna offers a client-server based environment in which a
Taverna \emph{server} can remotely serve the workflow composition and execution
services. This mode is useful in the cloud computing environments where the
server decides where to run the Taverna services in the cloud.

\paragraph{MOTEUR/scufl, GWENDIA} MOTEUR \cite{glatard-montagnat-etal:2008} is a SWE
developed at the I3S laboratory, University of Nice at Sophia Antipolis (UNS).
MOTEUR facilitates executing workflows consisting of remote services. It
exploits service parallelism at workflow level and data parallelism for
multiple datasets.

MOTEUR has sophisticated language-level expressions for data management
mechanisms within a workflow definition. It provides a diverse set of
parallel iterators that are well adapted for different combinations of input datasets
at the processor. An iteration strategy allows different combinations of the
data arriving at a processor along its multiple inputs. Running legacy
application code on distributed systems has been eased by MOTEUR by providing a
generic grid-service wrapper for these applications
\cite{glatard-montagnat-etal:2008a}. Services thus created can be readily
deployed by the MOTEUR runtime. MOTEUR workflows are written in the scufl
language and are compatible with those written for the Taverna SWE. It also
supports an extension of scufl language called the Gwendia~\cite{gwendia}
language, developed at the same laboratory. 
%
\paragraph{Makeflow/make-like spec} Makeflow~\cite{makeflow} is a make-like workflow platform.
Makeflow addresses the often-encountered problem of software and library
dependencies on workflow tasks execution for heterogeneous environments.
Makeflow offers techniques to ``link" and package the application execution
elements such as executables, libraries and configuration parameters. This
improves the durability of a workflow and decreases the efforts required to
maintain a working version of an application encoded as a workflow.

%Internal representation of a Makeflow workflow is in...
%Its expressivity features include...
%Makeflow's runtime features include...
%
\paragraph{Triana/XML} Triana \cite{taylor-shields-etal:2003} is a java-based
SWE developed at the Cardiff University. It consists of a workflow composition
GUI, a framework to integrate workflow with the different execution
environments and middlewares and a set of predefined services written as java
components. A Triana workflow is expressed as an XML grammar similar to a WSDL
description for webservices \cite{taylor-wand-etal:2005}. A highlight of
Triana is that it provides ready-to-use interface for a wide variety of
middleware and distributed infrastructures including P2P, grids and
web-services frameworks. A Triana workflow provides dataflow as well as control
flow support. Triana allows conditional branching and looping in its workflows.

%
\paragraph{Kepler/MoML} Kepler~\cite{kepler2013} is a SWE
based upon the Ptollemy II framework~\cite{eker-janneck-etal:2003} developed at
the University of California. Kepler workflows follow an actor-director
metaphor where an actor represents a processor and the director
represents the paradigm for executing the workflows consisting of the actors.
Depending upon the types of directors, a Kepler workflow may exhibit a
different type of execution behavior. A workflow designer can chose from among
directors based upon the purpose of a workflow. Few of these include Process
Network (PN) for a parallel execution at the workflow level, Synchronous Data
Flow (SDF) for simple sequential data flow and Collection Oriented Modeling and
Design (COMAD) \cite{keplercomad} for a collection oriented workflow. Kepler
uses the Modeling Markup Language (MoML) for its workflow representation. MoML
is an extended XML syntax providing a rich vocabulary to compose Kepler
workflows. A workflow processor is called an `actor' in Kepler. Recent advances
in Kepler include MapReduce style computation, bioinformatics computation,
cloud computing~\cite{kepler2013}.

%
\paragraph{Pegasus/DAX} Pegasus \cite{deelman-peg} is an SWE
well-suited to generate and execute large workflows from abstract DAGs
describing dataflow among interdependent tasks. Abstract description of a DAG
is done via an XML-based language called DAX. The concrete expression takes
into account the properties of application task requirements and underlying
distributed compute resources.  The main advantage of such approach is that a
single abstract workflow description can be mapped to multiple execution
environments based on requirements. DAX describes the logical workflow
components, their inputs and outputs and their links and dependencies in a
Pegasus workflow. These kind of workflows have been shown to enact optimally
using Pegasus' dynamic task, memory and disk management techniques. For
instance, for massively parallel task scheduling, Pegasus provides a task
clustering technique that groups parallel jobs based on fixed parameters such
as number of groups. Additionally, a manager buffers and maintains a fix-sized
window of executable tasks before accepting new tasks.

Large scale workflows from astronomy, biology, and high-energy physics have
been shown to run via Pegasus on distributed computing resources such as
Clusters, Grids and Supercomputers.  Recent developments in Pegasus enables it
to run workflows on cloud environments~\cite{peg-cloud}. Pegasus system is
known to be extendible as illustrated by a recent integration
effort~\cite{peg-hubzero} between Pegasus and the Hubzero
portal~\cite{hubzero}.

\paragraph{Wings/--} The Workflow instance generation and specialization
(Wings) provides automatic workflow validation and generation based on
high-level workflow templates~\cite{wings2013}. The system receives high-level
workflow specification in the form of Pegasus/DAX workflows and validates them
for correct coordination based on data and resource constraints. It generates
workflows by selecting components, data sets, and parameters in three stages
while eliminating candidates that are not viable. This approach is useful in
the context of desiging complex computational experiments by domain scientists
in that it eases the process of workflow level debugging. Wings uses W3C's OWL,
Resource Description Framework (RDF), and Semantic Web Rule Language (SWRL) to
represent workflows and their associated constraints.

% How does Wings achieve the translation between abstract and concrete workflows?
% What are the constraints under which Wings works?
% How is Wings implemented?

%
\paragraph{Swift/Script, Tcl} Swift~\cite{swift} is a script-based scientific
workflow environment developed at the University of Chicago and Argonne
National Laboratory. A Swift workflow specified in the Swift scripting language
and is translated to an intermediate XML-based representation and enacted by
the Karajan workflow enactment engine. The SwiftScript language is proposed as
a workflow language founded on a scripting approach that programmers are
familiar with. To make the language data parallel, Swift script generalizes the
use of futures~\cite{futures} for every variables defined in the
script.  Futures are non-blocking assignment variables, using a proxy to ensure
immediate execution of assignment and performing lazy blocking on variable
value accesses only. The systematic use of futures for every variables in the
language makes it completely data-driven: the execution progresses
asynchronously as long as a data access is not blocking. The availability of
data enables blocked thread to restart execution. Swift facilitates a highly
asynchronous workflow execution. Swift supports advanced workflow constructs
such as parallel loops and conditionals. Furthermore Swift supports processing
of explicitly defined arrays of data. Swift workflows take advantage of
POSIX-style shared file system to reuse the data produced by other processors.
Recently, Swift has been shown to run on multiple clouds~\cite{ccgrid2013} via
its coasters mechanism~\cite{ucc2011}. Swift/T~\cite{turbine2013}, a new HPC
implementation of Swift system, runs scientific workflows over large-scale
supercomputing resources. It has been shown to work on petascale supercomputers
such as the IBM BlueGene and Cray series of machines. GeMTC project
investigating running many-task applications over GPUs via Swift/T. Efforts to develop metadata and provenance modules have been undertaken recently~\cite{swiftprov}.

% Put this somewhere else, either in the analysis or may be in the conclusions
% and summary.
%The \textit{gscript}~\cite{maheshwari-montagnat:2010} workflow language
%is similar in spirit to the
%SwiftScript. Similar to SwiftScript, gscript adopts futures to express
%asynchronous execution in a script-based language. A Swift script 
%contains statically typed datasets that can be declared
%in the script. Unlike Swift script, gscript does provide user-defined
%strategies in order to effect data combinations in the workflow. The underlying datasets organization in Swift is
%represented using XML-representation. This means that effectively, a
%workflow is a DAG. A workflow is
%simpler and compact in gscript as its data description is embedded within a
%workflow representation. Conditional and Loop execution is supported
%using the if and foreach statements in both languages.
%

\paragraph{ASKALON/AGWL.} The ASKALON Problem Solving Environment (PSE)
\cite{fahringer2014} is developed at the University of Innsbruck.
ASKALON is described as a Scientific PSE capable of composing, hosting,
executing, sharing and reusing scientific workflows written in the Abstract
Grid Workflow Language (AGWL). AGWL is an XML-based workflow language that is
capable of expressing scientific application components chains at a high-level
of abstraction. A concrete workflow consists of the abstract workflow combined
with the resources and target execution environment binding the workflow. The
graphical interface of AGWL is based upon the UML (Unified Modeling Language)
diagrams.\km{Todo: current developments and advances.}

\paragraph{The P-GRADE Portal/--} The P-GRADE Portal~\cite{pgrade2011}
developed at the MTA SZTAKI Computation and Automation Institute, Budapest,
provides an access to grid computational and storage resources from a web
browser through a specialized workflow oriented grid portal. A workflow
development environment interfaces the users through this web based portal. The
portal interfaces the multiple grid computing middleware and storage resources
as a backend component of the environment. A highlight of P-GRADE portal is
that it provides access to multiple grids and scientific communities in a
single environment leading to a highly collaborative inter-community,
inter-grid research environment. An older version of P-GRADE used the
condor-Dagman as workflow engine while the recent version uses its own in-house
developed workflow engine called Xen. \km{Investigate cloud and GPU capabilities} 
%

\paragraph{Galaxy Portal/JSON} Galaxy~\cite{blankenberg-vonkuster-etal:2010} is a
SWE developed at the Penn State University, Pennsilvania. It provides a
webportal based environment to build complex pipelines from an available
repository of tools (\textit{http://g2.bx.psu.edu}).  Galaxy is currently used
mainly by bioinformatics community. Users can extend the tool repository
provided by Galaxy using their own Galaxy server. One of the major use of
Galaxy is to achieve a mix and match of data analysis tools in order to analyze
a vast repository of genome based database. Galaxy offers a web-based,
interactive platform for data analysis via reusable and reproducible workflows.
While excelling in the usability aspects, the Galaxy environment has a limited
parallelism capabilities and a small number of existing implementations for
interfacing with diverse, scalable computing infrastructures. The user can draw
workflows in which multiple copies of a task are created to operate on multiple
inputs at once, but this is not a scalable approach.

>>>>

\input{text/taverna}
\input{text/moteur}
\input{text/makeflow}
\input{text/triana}
\input{text/kepler}
\input{text/peg}
\input{text/wings}
\input{text/swift}
\input{text/askalon}
\input{text/pgrade}
\input{text/galaxy}
% Put this somewhere else, either in the analysis or may be in the conclusions
% and summary.
%The \textit{gscript}~\cite{maheshwari-montagnat:2010} workflow language
%is similar in spirit to the
%SwiftScript. Similar to SwiftScript, gscript adopts futures to express
%asynchronous execution in a script-based language. A Swift script 
%contains statically typed datasets that can be declared
%in the script. Unlike Swift script, gscript does provide user-defined
%strategies in order to effect data combinations in the workflow. The underlying datasets organization in Swift is
%represented using XML-representation. This means that effectively, a
%workflow is a DAG. A workflow is
%simpler and compact in gscript as its data description is embedded within a
%workflow representation. Conditional and Loop execution is supported
%using the if and foreach statements in both languages.
%
%
